---
title: "Secretarybird_MonthlyMovement_AutomatedReport"
author: "C.W. Brink"
date: "2022-11-14"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Secretarybird Project Background

BirdLife South Africa’s Secretarybird Project started in 2011, with the initial main aim of obtaining movement data for immature and adult birds. The first phase of the project, to fit tracking devices to juvenile birds, was successfully completed (2012-2015) and the first analyses from these data have been written as a scientific article for submission in early July 2018 to Ostrich-Journal of African Ornithology.

The second phase of this study investigates the habitat requirements of Secretarybirds and differences in birds foraging in different habitats. To enable these additional analysis additional Secretarybirds have been added to the tracking study. Additionally, there are future plans to investigate the future effects of Climate Change on this species.

The information garnered from this tracking study contribute to the main aim of the Secretarybird Project which is to identify and implement priority conservation actions for Secretarybirds and larger ecosystems. Effective conservation necessitates actions that are based on scientific evidence and research is thus an important component of this project.

BirdLife South Africa recognizes the potential to use Secretarybirds as a flagship species to promote an interest in conservation and the correct land management practices required to benefit other bird of prey species, as well as a suite of threatened, endemic and common bird species.


This document gives a basic overview of the movements of Secretarybirds tracked by BirdLife South Africa as part of their ongoing movement ecology study, up until the date: `r format(Sys.time(), '%Y/%m/%d')`. Please note that all distance measurements are the straight line distance calculated between points and are likely and underestimation of the distances covered by each bird given the nature of how Secretarybirds move.


```{r Load Packages}

rm(list=ls())

library(RCurl)
library(bitops)
library(move)
library(mapdata)
library(ggmap)
library(mapproj)
library(dplyr)
library(sf)
library(lubridate)
library(here)

options(scipen=999)

see <- function(x) utils::View(x)

cred <- movebankLogin(username = "xxx", password = "xxx")

sb <- getMovebankData(study = "Secretarybird Sagittariidae South Africa", login = cred,
                      removeDuplicatedTimestamps=TRUE
                      ) # all data

```

```{r Data Cleaning}
# needs to be df not move object for this
sb.df <- as.data.frame(sb)
# checking for timestamp duplicates - seems there are none
any(duplicated(sb.df[,c("timestamp", "local_identifier")])) 
# removing duplicates
reduced <- sb.df[!duplicated(
    sb.df[, c("timestamp", "location_long", "location_lat", "local_identifier")]), ]
# accounting for multi-locations:
# first ensure that timestamps and individuals are ordered this will
# make sure the previous record is the same when checking for duplicates
reduced <- reduced[order(reduced$local_identifier, reduced$timestamps),]
# identify first duplicate
dup <- anyDuplicated(reduced[,c('timestamp', 'local_identifier')])


# THERE ARE NO DUPLICATES - if fuplicates are found refer back to source material for cleaning

# define the data.frame as a move object after cleaning
secb <- move(x=reduced$location_long,
  y=reduced$location_lat,
  time=reduced$timestamp,
  data=reduced,
  proj=CRS("+proj=longlat +datum=WGS84"),
  animal=reduced$nick_name)

```

```{r All birds}
(e <- bbox(extent(secb)*2.8))

tiff(here::here("Outputs", "AllTracks.tiff"), units="cm", width=20, height=15, res=300)
map('worldHires', xlim = e[1, ], ylim = e[2, ])
points(secb)
dev.off()

```

```{r All birds plot, out.width = "85%", fig.cap = "Figure 1: Movements of all Secretarybird tracked by BirdLife South Africa since October 2020."}
map('worldHires', xlim = e[1, ], ylim = e[2, ])
points(secb)
```


## Ekapa

```{r Ekapa Data import}

bird <- secb[["Ekapa"]]

```

```{r Ekapa Movement Map}

(e <- bbox(extent(bird)*1.5)) # gives indication of coordinate extent of ekapa tracks (but multiplied by factor of 5)

# ggplot map:

bird_df <- as(bird, "data.frame") # this google method requires df

m <- get_map(e, zoom = 9, source = "google", maptype = "terrain")

bird_map <- ggmap(m) + geom_path(data = bird_df, aes(x = location_long, y = location_lat, colour = "red")) + 
  theme(legend.position = "none")

# ekap <- data[with(data, order(UTC_datetime)), ] 

```

```{r Ekapa Database prep}

bird_df_simple <- bird_df[ , c("timestamp",  "timestamps", "location_lat", "location_long")]

CRS("+init=epsg:32733")

bird_sf <- bird_df_simple %>%
sf::st_as_sf(coords = c("location_long", "location_lat"), crs = 4326) %>%
sf::st_transform(crs = 32733)

```

```{r Ekapa Distance between points}

Point_dist <- bird_sf %>% 
  mutate(
    lead = geometry[row_number()+1],
    dist = st_distance(geometry, lead, by_element = T) # this code does what I need - build on it
  )

Dist_per_month <- Point_dist %>% 
    group_by(month = lubridate::ceiling_date(timestamp, "month")) %>%
    summarize(summary_variable = sum(dist))

Dist_per_month <- na.omit(Dist_per_month)

Dist_per_month$km <- as.numeric(Dist_per_month$summary_variable/1000)

# Make Graph

Dist_per_month$month <- as.Date(Dist_per_month$month)

library(scales)
Bird_movemet_sum <- ggplot(Dist_per_month, aes(x = month, y = km)) +
  geom_bar(stat = "identity") +
  ylab("Distance traveled (km)") +
  scale_y_continuous(breaks=seq(0, max(Dist_per_month$km), by=100))+
  scale_x_date(date_breaks = "1 month", date_labels = "%B %y") + #library(scales) is needed
  theme(axis.title.x=element_blank(), 
        axis.text.x = element_text(angle = 70, vjust = 1, hjust = 1),
        panel.background = element_rect(fill = "transparent",colour = NA),
        axis.line = element_line(size = 0.5, colour = "black", linetype = 1))

```

Ekapa is a male Secretarybird whose device was fitted while he was still a chick on his nest on 2020/10/22 near Kimberley, Northern Cape. This device was sponsored by Ekapa Minerals through support from Ester van der Westhuizen. Since the deployment of his device Ekapa has traveled roughly `r format(round(sum(Dist_per_month$km), 2))` km (Figure 2 & 3).

```{r ekapa_map, out.width = "100%", fig.cap = "Figure 2: Movements of the Secretarybird, Ekapa, since the deployment of his device at his nest site near Kimberley, Northern Cape.", resutls = 'markup'}

bird_map

```

```{r, fig.cap = "Figure 3: Monthly total distances traveled by Ekapa since his device deployment.", resutls = 'markup'}

Bird_movemet_sum

```

```{r}
tiff(here::here("Outputs", "EkapaTracks.tiff"), units="cm", width=20, height=15, res=300)
bird_map
dev.off()
```

```{r}
tiff(here::here("Outputs", "Ekapa_graph.tiff"), units="cm", width=20, height=15, res=300)
bird_map
dev.off()
```
 
## Egoli

```{r Egoli Data import}

bird <- secb[["Egoli"]]

```

```{r Egoli Movement Map}

(e <- bbox(extent(bird)*1.5)) # gives indication of coordinate extent of ekapa tracks (but multiplied by factor of 5)

# ggplot map:

bird_df <- as(bird, "data.frame") # this google method requires df

m <- get_map(e, zoom = 9, source = "google", maptype = "terrain")

bird_map <- ggmap(m) + geom_path(data = bird_df, aes(x = location_long, y = location_lat, colour = "red")) + 
  theme(legend.position = "none")

```

```{r Egoli Database prep}

bird_df_simple <- bird_df[ , c("timestamp",  "timestamps", "location_lat", "location_long")]

CRS("+init=epsg:32733")

bird_sf <- bird_df_simple %>%
sf::st_as_sf(coords = c("location_long", "location_lat"), crs = 4326) %>%
sf::st_transform(crs = 32733)

```

```{r Egoli Distance between points}

Point_dist <- bird_sf %>% 
  mutate(
    lead = geometry[row_number()+1],
    dist = st_distance(geometry, lead, by_element = T) # this code does what I need - build on it
  )

Dist_per_month <- Point_dist %>% 
    group_by(month = lubridate::ceiling_date(timestamp, "month")) %>%
    summarize(summary_variable = sum(dist))

Dist_per_month <- na.omit(Dist_per_month)

Dist_per_month$km <- as.numeric(Dist_per_month$summary_variable/1000)

# Make Graph

Dist_per_month$month <- as.Date(Dist_per_month$month)

library(scales)
Bird_movemet_sum <- ggplot(Dist_per_month, aes(x = month, y = km)) +
  geom_bar(stat = "identity") +
  ylab("Distance traveled (km)") +
  scale_y_continuous(breaks=seq(0, max(Dist_per_month$km), by=100))+
  scale_x_date(date_breaks = "1 month", date_labels = "%B %y") + #library(scales) is needed
  theme(axis.title.x=element_blank(), 
        axis.text.x = element_text(angle = 70, vjust = 1, hjust = 1),
        panel.background = element_rect(fill = "transparent",colour = NA),
        axis.line = element_line(size = 0.5, colour = "black", linetype = 1))

```

Egoli is a male Secretarybird whose device was fitted while he was still a chick on his nest on 2021/09/09 near Heidelberg, Gauteng. This device was sponsored by BirdLife Northern Gauteng. Egoli is the sibling from the same brood as Mia-May (see below). Since the deployment of his device Egoli has traveled roughly `r format(round(sum(Dist_per_month$km), 2))` km (Figure 4 & 5).

```{r egoli_map, out.width = "85%", fig.cap = "Figure 4: Movements of the Secretarybird, Egoli, since the deployment of his device at his nest site near Heidelberg, Gauteng.", resutls = 'markup'}

bird_map

```

```{r, fig.cap = "Figure 5: Monthly total distances traveled by Egoli since his device deployment.", resutls = 'markup'}

Bird_movemet_sum

```

## Mia-May

```{r MiaMay Data import}

bird <- secb[["MiaMay"]]

```

```{r MiaMay Movement Map}

(e <- bbox(extent(bird)*1.5)) # gives indication of coordinate extent of ekapa tracks (but multiplied by factor of 5)

# ggplot map:

bird_df <- as(bird, "data.frame") # this google method requires df

m <- get_map(e, zoom = 9, source = "google", maptype = "terrain")

bird_map <- ggmap(m) + geom_path(data = bird_df, aes(x = location_long, y = location_lat, colour = "red")) + 
  theme(legend.position = "none")

```

```{r MiaMay Database prep}

bird_df_simple <- bird_df[ , c("timestamp",  "timestamps", "location_lat", "location_long")]

CRS("+init=epsg:32733")

bird_sf <- bird_df_simple %>%
sf::st_as_sf(coords = c("location_long", "location_lat"), crs = 4326) %>%
sf::st_transform(crs = 32733)

```

```{r MiaMay Distance between points}

Point_dist <- bird_sf %>% 
  mutate(
    lead = geometry[row_number()+1],
    dist = st_distance(geometry, lead, by_element = T) # this code does what I need - build on it
  )

Dist_per_month <- Point_dist %>% 
    group_by(month = lubridate::ceiling_date(timestamp, "month")) %>%
    summarize(summary_variable = sum(dist))

Dist_per_month <- na.omit(Dist_per_month)

Dist_per_month$km <- as.numeric(Dist_per_month$summary_variable/1000)

# Make Graph

Dist_per_month$month <- as.Date(Dist_per_month$month)

library(scales)
Bird_movemet_sum <- ggplot(Dist_per_month, aes(x = month, y = km)) +
  geom_bar(stat = "identity") +
  ylab("Distance traveled (km)") +
  scale_y_continuous(breaks=seq(0, max(Dist_per_month$km), by=100))+
  scale_x_date(date_breaks = "1 month", date_labels = "%B %y") + #library(scales) is needed
  theme(axis.title.x=element_blank(), 
        axis.text.x = element_text(angle = 70, vjust = 1, hjust = 1),
        panel.background = element_rect(fill = "transparent",colour = NA),
        axis.line = element_line(size = 0.5, colour = "black", linetype = 1))

```

MiaMay is a female Secretarybird whose device was fitted while she was still a chick on her nest on 2021/09/09 near Heidelberg, Gauteng. Mia-May is the brood sibling to Egoli. Since the deployment of her device Mia-May has traveled roughly `r format(round(sum(Dist_per_month$km), 2))` km (Figure 6 & 7).

```{r MiaMay_map, out.width = "85%", fig.cap = "Figure 6: Movements of the Secretarybird, MiaMay, since the deployment of her device at her nest site near Heidelberg, Gauteng.", resutls = 'markup'}

bird_map

```

```{r, fig.cap = "Figure 7: Monthly total distances traveled by Mia-May since her device deployment.", resutls = 'markup'}

Bird_movemet_sum

```


## Fawkes

```{r Fawkes Data import}

bird <- secb[["Fawkes"]]

```

```{r Fawkes Movement Map}

(e <- bbox(extent(bird)*1.5)) # gives indication of coordinate extent of ekapa tracks (but multiplied by factor of 5)

# ggplot map:

bird_df <- as(bird, "data.frame") # this google method requires df

m <- get_map(e, zoom = 9, source = "google", maptype = "terrain")

bird_map <- ggmap(m) + geom_path(data = bird_df, aes(x = location_long, y = location_lat, colour = "red")) + 
  theme(legend.position = "none")

```

```{r Fawkes Database prep}

bird_df_simple <- bird_df[ , c("timestamp",  "timestamps", "location_lat", "location_long")]

CRS("+init=epsg:32733")

bird_sf <- bird_df_simple %>%
sf::st_as_sf(coords = c("location_long", "location_lat"), crs = 4326) %>%
sf::st_transform(crs = 32733)

```

```{r Fawkes Distance between points}

Point_dist <- bird_sf %>% 
  mutate(
    lead = geometry[row_number()+1],
    dist = st_distance(geometry, lead, by_element = T) # this code does what I need - build on it
  )

Dist_per_month <- Point_dist %>% 
    group_by(month = lubridate::ceiling_date(timestamp, "month")) %>%
    summarize(summary_variable = sum(dist))

Dist_per_month <- na.omit(Dist_per_month)

Dist_per_month$km <- as.numeric(Dist_per_month$summary_variable/1000)

# Make Graph

Dist_per_month$month <- as.Date(Dist_per_month$month)

library(scales)
Bird_movemet_sum <- ggplot(Dist_per_month, aes(x = month, y = km)) +
  geom_bar(stat = "identity") +
  ylab("Distance traveled (km)") +
  scale_y_continuous(breaks=seq(0, max(Dist_per_month$km), by=100))+
  scale_x_date(date_breaks = "1 month", date_labels = "%B %y") + #library(scales) is needed
  theme(axis.title.x=element_blank(), 
        axis.text.x = element_text(angle = 70, vjust = 1, hjust = 1),
        panel.background = element_rect(fill = "transparent",colour = NA),
        axis.line = element_line(size = 0.5, colour = "black", linetype = 1))

```

Fawkes is a male Secretarybird whose device was fitted while he was still a chick on his nest on 2021/12/21 at Babanango Game Reserve, KwaZulu-Natal. Since the deployment of his device Fawkes has traveled roughly `r format(round(sum(Dist_per_month$km), 2))` km (Figure 8 & 9).

```{r Fawkes_map, out.width = "85%", fig.cap = "Figure 8: Movements of the Secretarybird, Fawkes, since the deployment of his device at his nest site in Babanango Nature Reserve, KwaZulu-Natal.", resutls = 'markup'}

bird_map

```

```{r, fig.cap = "Figure 9: Monthly total distances traveled by Fawkes since his device deployment.", resutls = 'markup'}

Bird_movemet_sum

```


## Bob

```{r Bob Data import}

bird <- secb[["Bob"]]

```

```{r Bob Movement Map}

(e <- bbox(extent(bird)*1.5)) # gives indication of coordinate extent of ekapa tracks (but multiplied by factor of 5)

# ggplot map:

bird_df <- as(bird, "data.frame") # this google method requires df

m <- get_map(e, zoom = 9, source = "google", maptype = "terrain")

bird_map <- ggmap(m) + geom_path(data = bird_df, aes(x = location_long, y = location_lat, colour = "red")) + 
  theme(legend.position = "none")

```

```{r Bob Database prep}

bird_df_simple <- bird_df[ , c("timestamp",  "timestamps", "location_lat", "location_long")]

CRS("+init=epsg:32733")

bird_sf <- bird_df_simple %>%
sf::st_as_sf(coords = c("location_long", "location_lat"), crs = 4326) %>%
sf::st_transform(crs = 32733)

```

```{r Bob Distance between points}

Point_dist <- bird_sf %>% 
  mutate(
    lead = geometry[row_number()+1],
    dist = st_distance(geometry, lead, by_element = T) # this code does what I need - build on it
  )

Dist_per_month <- Point_dist %>% 
    group_by(month = lubridate::ceiling_date(timestamp, "month")) %>%
    summarize(summary_variable = sum(dist))

Dist_per_month <- na.omit(Dist_per_month)

Dist_per_month$km <- as.numeric(Dist_per_month$summary_variable/1000)

# Make Graph

Dist_per_month$month <- as.Date(Dist_per_month$month)

library(scales)
Bird_movemet_sum <- ggplot(Dist_per_month, aes(x = month, y = km)) +
  geom_bar(stat = "identity") +
  ylab("Distance traveled (km)") +
  scale_y_continuous(breaks=seq(0, max(Dist_per_month$km), by=100))+
  scale_x_date(date_breaks = "1 month", date_labels = "%B %y") + #library(scales) is needed
  theme(axis.title.x=element_blank(), 
        axis.text.x = element_text(angle = 70, vjust = 1, hjust = 1),
        panel.background = element_rect(fill = "transparent",colour = NA),
        axis.line = element_line(size = 0.5, colour = "black", linetype = 1))

```

Bob is a male Secretarybird whose device was fitted while he was still a chick on his nest on 2021/12/21 at Babanango Game Reserve, KwaZulu-Natal. Since the deployment of his device Bob has traveled roughly `r format(round(sum(Dist_per_month$km), 2))` km (Figure 10 & 11).

```{r Bob_map, out.width = "85%", fig.cap = "Figure 10: Movements of the Secretarybird, Bob, since the deployment of his device at his nest site in Babanango Nature Reserve, KwaZulu-Natal.", resutls = 'markup'}

bird_map

```

```{r, fig.cap = "Figure 11: Monthly total distances traveled by Bob since his device deployment.", resutls = 'markup'}

Bird_movemet_sum

```

## Nico

```{r Nico Data import}

bird <- secb[["Nico"]]

```

```{r Nico Movement Map}

(e <- bbox(extent(bird)*1.5)) # gives indication of coordinate extent of ekapa tracks (but multiplied by factor of 5)

# ggplot map:

bird_df <- as(bird, "data.frame") # this google method requires df

m <- get_map(e, zoom = 9, source = "google", maptype = "terrain")

bird_map <- ggmap(m) + geom_path(data = bird_df, aes(x = location_long, y = location_lat, colour = "red")) + 
  theme(legend.position = "none")

```

```{r Nico Database prep}

bird_df_simple <- bird_df[ , c("timestamp",  "timestamps", "location_lat", "location_long")]

CRS("+init=epsg:32733")

bird_sf <- bird_df_simple %>%
sf::st_as_sf(coords = c("location_long", "location_lat"), crs = 4326) %>%
sf::st_transform(crs = 32733)

```

```{r Nico Distance between points}

Point_dist <- bird_sf %>% 
  mutate(
    lead = geometry[row_number()+1],
    dist = st_distance(geometry, lead, by_element = T) # this code does what I need - build on it
  )

Dist_per_month <- Point_dist %>% 
    group_by(month = lubridate::ceiling_date(timestamp, "month")) %>%
    summarize(summary_variable = sum(dist))

Dist_per_month <- na.omit(Dist_per_month)

Dist_per_month$km <- as.numeric(Dist_per_month$summary_variable/1000)

# Make Graph

Dist_per_month$month <- as.Date(Dist_per_month$month)

library(scales)
Bird_movemet_sum <- ggplot(Dist_per_month, aes(x = month, y = km)) +
  geom_bar(stat = "identity") +
  ylab("Distance traveled (km)") +
  scale_y_continuous(breaks=seq(0, max(Dist_per_month$km), by=100))+
  scale_x_date(date_breaks = "1 month", date_labels = "%B %y") + #library(scales) is needed
  theme(axis.title.x=element_blank(), 
        axis.text.x = element_text(angle = 70, vjust = 1, hjust = 1),
        panel.background = element_rect(fill = "transparent",colour = NA),
        axis.line = element_line(size = 0.5, colour = "black", linetype = 1))

```

Nico is a male Secretarybird whose device was fitted while he was still a chick on his nest on 2021/12/16 in Randfontein, Gauteng. This device was sponsored by BirdLife Northern Gauteng. Since the deployment of his device Nico has traveled roughly `r format(round(sum(Dist_per_month$km), 2))` km (Figure 12 & 13).

```{r Nico_map, out.width = "85%", fig.cap = "Figure 12: Movements of the Secretarybird, Nico, since the deployment of his device at his nest site near Randfontein, Gauteng.", resutls = 'markup'}

bird_map

```

```{r, fig.cap = "Figure 13: Monthly total distances traveled by Nico since his device deployment.", resutls = 'markup'}

Bird_movemet_sum

```


## BLiNG2

```{r BLiNG2 Data import}

bird <- secb[["BLiNG2"]]

```

```{r BLiNG2 Movement Map}

(e <- bbox(extent(bird)*1.5)) # gives indication of coordinate extent of ekapa tracks (but multiplied by factor of 5)

# ggplot map:

bird_df <- as(bird, "data.frame") # this google method requires df

m <- get_map(e, zoom = 9, source = "google", maptype = "terrain")

bird_map <- ggmap(m) + geom_path(data = bird_df, aes(x = location_long, y = location_lat, colour = "red")) + 
  theme(legend.position = "none")

```

```{r BLiNG2 Database prep}

bird_df_simple <- bird_df[ , c("timestamp",  "timestamps", "location_lat", "location_long")]

CRS("+init=epsg:32733")

bird_sf <- bird_df_simple %>%
sf::st_as_sf(coords = c("location_long", "location_lat"), crs = 4326) %>%
sf::st_transform(crs = 32733)

```

```{r BLiNG2 Distance between points}

Point_dist <- bird_sf %>% 
  mutate(
    lead = geometry[row_number()+1],
    dist = st_distance(geometry, lead, by_element = T) # this code does what I need - build on it
  )

Dist_per_month <- Point_dist %>% 
    group_by(month = lubridate::ceiling_date(timestamp, "month")) %>%
    summarize(summary_variable = sum(dist))

Dist_per_month <- na.omit(Dist_per_month)

Dist_per_month$km <- as.numeric(Dist_per_month$summary_variable/1000)

# Make Graph

Dist_per_month$month <- as.Date(Dist_per_month$month)

library(scales)
Bird_movemet_sum <- ggplot(Dist_per_month, aes(x = month, y = km)) +
  geom_bar(stat = "identity") +
  ylab("Distance traveled (km)") +
  scale_y_continuous(breaks=seq(0, max(Dist_per_month$km), by=100))+
  scale_x_date(date_breaks = "1 month", date_labels = "%B %y") + #library(scales) is needed
  theme(axis.title.x=element_blank(), 
        axis.text.x = element_text(angle = 70, vjust = 1, hjust = 1),
        panel.background = element_rect(fill = "transparent",colour = NA),
        axis.line = element_line(size = 0.5, colour = "black", linetype = 1))

```

BLiNG2 is a male Secretarybird whose device was fitted while he was still a chick on his nest on 2021/12/21  on Sondela Nature Reserve near Bela-Bela, Limpopo. This device was sponsored by BirdLife Northern Gauteng. Since the deployment of his device BLiNG2 has traveled roughly `r format(round(sum(Dist_per_month$km), 2))` km (Figure 14 & 15).

```{r BLiNG2_map, out.width = "85%", fig.cap = "Figure 14: Movements of the Secretarybird, BLiNG2, since the deployment of his device at his nest site in Sondela Nature Reserve, Limpopo.", resutls = 'markup'}

bird_map

```

```{r, fig.cap = "Figure 15: Monthly total distances traveled by BLiNG2 since his device deployment.", resutls = 'markup'}

Bird_movemet_sum

```

We are grateful to the many partners and sponsors of the BirdLife South Africa Secretarybird Project including t . . . We are also grateful to the landowners, supporters and collaborators of this project including  . . . 
